Text-mining clinically relevant cancer biomarkers for curation into the CIViC database

### 2019-09-13 paper/paper.Rmd
Bigger changes to paper draft

---
title: Text-mining clinically relevant cancer biomarkers for curation into the CIViC database
#output: 
#  bookdown::pdf_document2
#  html_document: 
#    fig_caption: yes
#params:
#   region: east
#editor_options: 
#  chunk_output_type: console
site: bookdown::bookdown_site
documentclass: book
output:
#  bookdown::html_document2
  bookdown::word_document2
#  bookdown::pdf_document2
bibliography: civicmine.bibtex
csl: plos-computational-biology.csl
---

@@

```{r civicmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('civicmine/fig_overview.R')
#read_chunk('fig_annotationAndModel.R')
source('civicmine/tab_keywords.R')
source('civicmine/tab_interannotatoragreement.R')
source('civicmine/fig_prcurves.R')
source('civicmine/fig_comparison.R')
source('civicmine/fig_results.R')
source('civicmine/fig_evaluationresults.R')
invisible(dev.off())
```

## Abstract
Precision oncology involves analysis of individual cancer samples to understand the genes and pathways involved in the development and progression of a cancer. To improve patient care, knowledge of diagnostic, prognostic, predisposing and drug response markers is essential. Several knowledgebases have been created by different groups to collate evidence for these associations. These include the open-access Clinical Interpretation of Variants in Cancer (CIViC) knowledgebase. These databases rely on time-consuming manual curation from skilled experts who read and interpret the relevant biomedical literature. To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant biomarkers from all available published literature. To this end, a group of cancer genomics experts annotated biomarkers and their clinical associations discussed in 800 sentences and achieved good inter-annotator agreement. We then used a supervised learning approach to construct the CIViCmine knowledgebase (http://bionlp.bcgsc.ca/civicmine/), extracting `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers. CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications. Through integration with CIVIC, we provide a prioritised list of curatable biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general.

@@

@@

IE methods extract structured knowledge directly from the papers. This can take the form of entity recognition, by explicitly tagging mentions of biomedical concepts such as genes, drugs and diseases. A further step can involve relation extraction to understand the relationship discussed between tagged biomedical entities. This structured information can then be used to identify papers relevant for the knowledgebase. IE methods are also used for automated knowledgebase population without a manual curation step. For example, the miRTex knowledgebase, which collates microRNAs and their targets, uses automated relation extraction methods to populate the knowledgebase [@li2015mirtex]. Protein-protein interaction networks (such as STRING [@szklarczyk2016string]) are often built using automatically generated knowledgebases.

The main objective of this project was to identify frequently discussed cancer biomarkers that fit the CIViC evidence model but are not yet included in the CIViC knowledgebase. We developed an IE-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full text articles and compare against the CIViC knowledgebase. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). The main contributions of this work are an approach for knowledgebase construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically-accessible set of abstracts and full text articles. This resource, containing `r paper.biomarkerCount` gene-cancer associations with clinical relevance, is valuable to all cancer knowledgebases to aid their curation and also as a tool for precision cancer analysts searching for evidence supporting biomarkers not yet included in any other resource.

@@

#keywords %>%
#  knitr::kable(
#    "html",
#    booktabs = TRUE, escape=FALSE,
#    caption = 'The five groups of filter strings  used to identify sentences that potentially discussed the four evidence types. Strings such as "sensitiv" are used to capture multiple words including "sensitive" and "sensitivity".'
#    ) %>%
#  kable_styling()

knitr::kable(
    keywords,
    booktabs = TRUE, escape=FALSE,
    caption = 'The five groups of search terms used to identify sentences that potentially discussed the four evidence types  . Strings such as "sensitiv" are used to capture multiple words including "sensitive" and "sensitivity".'
    )
```

@@

knitr::kable(
  traintest,
  booktabs = TRUE, 
  caption = 'Number of annotations in the training and test sets',
  col.names = c('Annotation','Train', 'Test')
  )
```

@@

```{r performance, echo=F, eval=T}
knitr::kable(
  paper.performanceTable,
  booktabs = TRUE, escape=FALSE,
  row.names=F,
  caption = 'The selected thresholds for each relation type with the high precision and lower recall trade-off.',
  col.names = c("Extracted Relation","Threshold","Precision","Recall")
  )
```

@@

suppressWarnings(exampleSentences %>%
  dplyr::mutate_all(linebreak) %>%
  knitr::kable(
    booktabs = TRUE, escape=FALSE,
    caption = 'Four example sentences for the four evidence types extracted by CIViCmine. The associated PubMed IDs are also shown for reference.',
    col.names = c('Type','PMID', 'Sentence'),
    row.names = NA
    ) %>%
  kable_styling())
```

@@

# Acknowledgements

@@

>>>>

---
title: Text-mining clinically relevant cancer biomarkers for curation into the CIViC database
#output: 
#  bookdown::pdf_document2
#  html_document: 
#    fig_caption: yes
#params:
#   region: east
#editor_options: 
#  chunk_output_type: console
site: bookdown::bookdown_site
documentclass: book
output:
#  bookdown::html_document2
  bookdown::word_document2
#  bookdown::pdf_document2
bibliography: civicmine.bibtex
csl: vancouver.csl
---

@@

```{r civicmineFigures, echo=F}
pdf(file="sideeffects.pdf")
source('civicmine/tab_keywords.R')
source('civicmine/tab_interannotatoragreement.R')
source('civicmine/fig_overview.R')
source('civicmine/fig_prcurves.R')
source('civicmine/fig_comparison.R')
source('civicmine/fig_results.R')
source('civicmine/fig_evaluationresults.R')
invisible(dev.off())

```

## Abstract

**Background**: Precision oncology involves analysis of individual cancer samples to understand the genes and pathways involved in the development and progression of a cancer. To improve patient care, knowledge of diagnostic, prognostic, predisposing and drug response markers is essential. Several knowledgebases have been created by different groups to collate evidence for these associations. These include the open-access Clinical Interpretation of Variants in Cancer (CIViC) knowledgebase. These databases rely on time-consuming manual curation from skilled experts who read and interpret the relevant biomedical literature.

**Methods**: To aid in this curation and provide the greatest coverage for these databases, particularly CIViC, we propose the use of text mining approaches to extract these clinically relevant biomarkers from all available published literature. To this end, a group of cancer genomics experts annotated biomarkers and their clinical associations discussed in 800 sentences and achieved good inter-annotator agreement. We then used a supervised learning approach to construct the CIViCmine knowledgebase (http://bionlp.bcgsc.ca/civicmine/)

**Results**: We extracted `r paper.sentenceCount` relevant sentences from PubMed abstracts and Pubmed Central Open Access full text papers. CIViCmine contains over `r paper.biomarkerCount` biomarkers associated with `r paper.geneCount` genes, `r paper.drugCount` drugs and `r paper.cancerCount` cancer types, representing `r paper.abstractCount` abstracts and `r paper.articleCount` full-text publications.

**Conclusions**: Through integration with CIVIC, we provide a prioritised list of curatable clinically relevant cancer biomarkers as well as a resource that is valuable to other knowledgebases and precision cancer analysts in general. All data is publically available and distributed with a Creative Commons Zero license.

## Keywords

Precision oncology, text mining, information extraction, machine learning, cancer biomarkers

@@

The main objective of this project was to identify frequently discussed cancer biomarkers that fit the CIViC evidence model but are not yet included in the CIViC knowledgebase. We developed an information extraction-based method to extract key parts of the evidence item: cancer type, gene, drug (where applicable) and the specific evidence type from published literature. This allows us to count the number of mentions of specific evidence items in abstracts and full text articles and compare against the CIViC knowledgebase. This paper will present our methods to develop this resource, known as CIViCmine (http://bionlp.bcgsc.ca/civicmine/). The main contributions of this work are an approach for knowledgebase construction that could be applied to many areas of biology and medicine, a machine learning method for extracting complicated relationships between four entity types, and extraction of relationships across the largest possible publically-accessible set of abstracts and full text articles. This resource, containing `r paper.biomarkerCount` gene-cancer associations with clinical relevance, is valuable to all cancer knowledgebases to aid their curation and also as a tool for precision cancer analysts searching for evidence supporting biomarkers not yet included in any other resource.

## Background

@@

IE methods extract structured knowledge directly from the papers. This can take the form of entity recognition, by explicitly tagging mentions of biomedical concepts such as genes, drugs and diseases. A further step can involve relation extraction to understand the relationship discussed between tagged biomedical entities. This structured information can then be used to identify papers relevant for the knowledgebase. IE methods are also used for automated knowledgebase population without a manual curation step. For example, the miRTex knowledgebase, which collates microRNAs and their targets, uses automated relation extraction methods to populate the knowledgebase [@li2015mirtex]. Protein-protein interaction networks (such as STRING [@szklarczyk2016string]) are often built using automatically generated knowledgebases. Our previous work has used information extraction methods to extract the role of genes in cancer but did not identify specific aberrations or the clinical relevance of them [@lever2019cancermine].

@@

knitr::kable(
    keywords, 'pandoc',
    booktabs = TRUE, escape=FALSE,
    caption = 'The five groups of search terms used to identify sentences that potentially discussed the four evidence types  . Strings such as "sensitiv" are used to capture multiple words including "sensitive" and "sensitivity".'
    )
```

@@

knitr::kable(
  traintest, 'pandoc',
  booktabs = TRUE, 
  caption = 'Number of annotations in the training and test sets',
  col.names = c('Annotation','Train', 'Test')
  )
```

@@

```{r performance, echo=F, eval=T}
knitr::kable(
  paper.performanceTable, 'pandoc',
  booktabs = TRUE, escape=FALSE,
  row.names=F,
  caption = 'The selected thresholds for each relation type with the high precision and lower recall trade-off.',
  col.names = c("Extracted Relation","Threshold","Precision","Recall")
  )
```

@@

exampleSentences %>%
  dplyr::mutate_all(linebreak) %>%
  knitr::kable( 'pandoc',
    booktabs = TRUE, escape=FALSE,
    caption = 'Four example sentences for the four evidence types extracted by CIViCmine. The associated PubMed IDs are also shown for reference.',
    col.names = c('Type','PMID', 'Sentence'),
    row.names = NA
    )
```

@@

# Declarations

## Ethics approval and consent to participate

Not applicable

## Consent for publication

Not applicable

## Availability of data and materials

The complete datasets are available in the Zenodo repository (https://doi.org/10.5281/zenodo.1472826) with the June 2019 release used for this paper (http://doi.org/10.5281/zenodo.3239054). The data can also be viewed through the web viewer (http://bionlp.bcgsc.ca/civicmine/) and subsets of the data can be downloaded there. The code for the analysis and web viewer is available at GitHub (https://github.com/jakelever/civicmine/).

## Competing interests

The authors declare that they have no competing interests.

## Funding

@@

## Author Contributions

JL, MRJ, OLG, MG and SJMJ developed the project idea. JL developed the methods, did the analyses and wrote the initial paper draft. JL, MRJ, AMD, KK, MB, JG, LC and MG annotated text data. OLG, MG and SJMJ supervised the project. All authors read and approved the final manuscript.

## Acknowledgements

The authors would like to thank Compute Canada for the use of compute infrastructure for this research.
